What We Eval in the Shadows

### 2021-08-13 paper/main.tex
Few fixes in 6

The R language was intended to be extensible. The combination of lazy
evaluation, \k{substitute} and \k{eval} are some of the tools given to
developers to this end. To understand \emph{why} developers use \eval, we
manually inspected top 117 sites which contribute to 90\% of the calls in the
corpus. This section gives some real-world examples, we simplify code for
clarity.

@@

\paragraph{Extracting Varargs}
\Eval is to obtain the source code of the arguments passed to a vararg
parameter. One example is a function that iterates over the contents of the
vararg until it finds a non-null value. \k{substitute(alist(...))} yields an
expression that represents a call to \k{alist} but where the \k{...} are
expended with the arguments supplied to the function. \Eval then returns a
\k{list} with unevaluated argument expressions which are evaluated sequentially
for their result. The reason \eval is needed is not to force
any more arguments once a non-null value was found.

@@

\paragraph{Controlled Evaluation}
Developer often want to control when and if expressions are evaluated. For
example, a package that provides users with the ability to make assertions and,
if they fail, produce legible error messages. The example here accepts a vararg
with a list of expressions. It extracts the associated code in the same way as
the extracting varags example and evaluates each of the expression sequentially
in the caller's environment. If an expression does not hold, an error is
generated.

@@

\paragraph{Shadowing}
%%% TODO ?????????????////
Many packages use \eval to implement variations of the \k{base} R functions.
For example, the \k{igraph} package provides the \k{do_call} function shown
below. This is a variant of the builtin \k{do.call} function with the
additional support for specifying the environment where the call should happen.
It constructs a call from a name and a list of arguments, and pass it to \eval.

@@

\paragraph{Boilerplate Removal} 
A variant of code generation is to let users write compact code that is
expended via \eval and \k{match.call}. The example uses \k{match.call} to
access the expression with which the function is called. The next three lines
construct a call to \k{stats::model.frame} with argument text of parameters
\k{'formula'}, \k{'data'}, and \k{'weights'}. Then, the call expression is
evaluated in the parent environment. The use of \k{match.call}, avoids repeated
use of \k{substitute} for extracting the argument expressions. It is extensible
as new parameter names can be easily added. This example illustrates one of the
most recurring usage of \eval that is used by the majority of packages that do
statistical modeling.

@@

\paragraph{Formula Resolution}
Many packages use \eval to evaluate formulas. The example extracts the
\k{"formula"} attribute attached to the \k{pdMat} object and evaluates it.

@@

\Eval allows functions to resolve symbols in different environments that the
current one. For example, the \k{data.table} redefines the subsetting
operator\footnote{It defines it as a triplet \texttt{[i, j, k]}, where expression
  \texttt{j} is used to calculate a subset of rows from \texttt{i} grouped by \texttt{k}.}
for data frames to enable compact queries. The following  
computes the average departure delays on 2014 Delta flights for each
origin airport.

@@

>>>>

The R language was intended to be extensible. The combination of lazy
evaluation, \k{substitute} and \k{eval} are some of the tools given to
developers to this end. To understand \emph{why} developers use \eval, we
manually inspected top 117 sites which contribute to 90\% of the calls in the
corpus. In this section we show 10 real-world examples form this batch. The
goal is to illustrate the variety of use cases that employ \eval in the R
ecosystem. The code snippets are simplified for clarity.

@@

\paragraph{Extracting Varargs}
\Eval can be used to obtain the source code of the arguments passed to a vararg
parameter. One example is a function that iterates over the contents of the
vararg until it finds a non-null value. \k{substitute(alist(...))} yields an
expression that represents a call to \k{alist} but where the \k{...} are
expended with the arguments supplied to the function. \Eval then returns a
\k{list} with unevaluated argument expressions which are evaluated sequentially
for their result. The reason \eval is needed is not to force
any more arguments once a non-null value was found.

@@

\paragraph{Controlled Evaluation}
\Eval can be used to control when and if expressions are evaluated. For
example, \k{assertthat} provides users with the ability to make assertions and,
if they fail, produce legible error messages. The example here accepts a vararg
with a list of expressions. It extracts the associated code in the same way
as the extracting varags example and evaluates each of the expression
sequentially in the caller's environment. If an expression does not hold, an
error is generated.

@@

\paragraph{Shadowing}
%%% TODO ?????????????//// <- left by Jan
Many packages use \eval to implement variations of the \k{base} R functions.
For example, the \k{igraph} package provides the \k{do_call} function shown
below. This is a variant of the builtin \k{do.call} function with the
additional support for specifying the environment where the call should happen.
It constructs a call from a name and a list of arguments, and pass it to \eval.

@@

\paragraph{Boilerplate Removal} 
A variant of code generation is to let users write compact code that is
expended via \eval and \k{match.call}. The example uses \k{match.call} to
access the expression with which the function is called. The next three lines
construct a call to \k{stats::model.frame} with argument text of parameters
\k{'formula'}, \k{'data'}, and \k{'weights'}. Then, the call expression is
evaluated in the parent environment. The use of \k{match.call}, avoids repeated
use of \k{substitute} for extracting the argument expressions. It is extensible
as new parameter names can be easily added. This example illustrates one of the
most recurring usage of \eval that is used by the majority of packages that do
statistical modeling, \ie the \c{model.frame} shape in Table~\ref{tab:minimizedexpressions}.

@@

% TODO: update to explain why the eval is needed or remove
\paragraph{Formula Resolution}
Many packages use \eval to evaluate formulas. The example extracts the
\k{"formula"} attribute attached to the \k{pdMat} object and evaluates it.

@@

\Eval allows functions to resolve symbols in different environments than the
current one. For example, the \k{data.table} redefines the subsetting
operator\footnote{It defines it as a triplet \texttt{[i, j, k]}, where expression
  \texttt{j} is used to calculate a subset of rows from \texttt{i} grouped by \texttt{k}.}
for data frames to enable compact queries. The following  
computes the average departure delays on 2014 Delta flights for each
origin airport.

@@

The \k{glue} package contains a variant that leverages the non-standard scoping
as well allowing to resolve interpolation in alternative environments. For
example, the following snippet pipes the subsetted data frame into the
\k{glue_data} function that uses it to resolve the symbols in string
interpolation.

\begin{minipage}{.95\textwidth}
\medskip
\begin{lstlisting}
> flights[carrier=="DL" & year == 2014, mean(dep_delay), by=origin] |> 
    glue_data("{origin} has average delay of {round(V1)} min")
# LGA has average delay of 11 min
# EWR has average delay of 15 min
# JFK has average delay of 12 min
\end{lstlisting}\medskip
\end{minipage}
